---
title: "Final Project-Group 10"
author: "Luke Richard, Carter Parks, Carter Wolf, Amanda Ohmer"
date: "2024-03-29"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

**National Vehicle Crash Data**


This data set shows the national crash data for motor vehicles from NHTSA. Each crash recorded has many columns of data, keyed to the "Fatality Analysis Report User Manual". We will first clean the data by picking a specific state to explore, in order to better relate traffic and set belt laws to the data. We expect seatbelt laws to decrease fatal crashes by percentage, but this will be the main purpose of the study.

We intend to study the effects of seatlbelt use in Iowa compared to traffic fatalities and traffic laws.

```{r}
library(dplyr)
vehicle75 <- read.csv("./FARS1975NationalCSV/VEHICLE.CSV")
vehicle80 <- read.csv("./FARS1980NationalCSV/VEHICLE.CSV")
vehicle85 <- read.csv("./FARS1985NationalCSV/VEHICLE.CSV")
vehicle90 <- read.csv("./FARS1990NationalCSV/VEHICLE.CSV")
vehicle95 <- read.csv("./FARS1995NationalCSV/VEHICLE.CSV")
vehicle00 <- read.csv("./FARS2000NationalCSV/VEHICLE.CSV")
vehicle05 <- read.csv("./FARS2005NationalCSV/VEHICLE.CSV")
vehicle10 <- read.csv("./FARS2010NationalCSV/VEHICLE.CSV")
vehicle15 <- read.csv("./FARS2015NationalCSV/vehicle.CSV")
vehicle20 <- read.csv("./FARS2020NationalCSV/vehicle.CSV")

vehicle75 <- vehicle75 %>% mutate(YEAR = 1975)
vehicle80 <- vehicle80 %>% mutate(YEAR = 1980)
vehicle85 <- vehicle85 %>% mutate(YEAR = 1985)
vehicle90 <- vehicle90 %>% mutate(YEAR = 1990)
vehicle95 <- vehicle95 %>% mutate(YEAR = 1995)
vehicle00 <- vehicle00 %>% mutate(YEAR = 2000)
vehicle05 <- vehicle05 %>% mutate(YEAR = 2005)
vehicle10 <- vehicle10 %>% mutate(YEAR = 2010)
vehicle15 <- vehicle15 %>% mutate(YEAR = 2015)
vehicle20 <- vehicle20 %>% mutate(YEAR = 2020)

vehicle75 <- vehicle75 %>%
  select(-CHAS_TR, -MCYCL_TY, -DR_TRAIN, -VIOL_CHG, -FLDCD_TR, -IMPACTS, -TOWAWAY, -VEH_CF1, -VEH_CF2, -DR_CF1, -DR_CF2, -DR_CF3, -IMPACT2, -WGTCD_TR, -VIN_LNGT, -VIN_WGT, -WHLBS_SH, -WHLBS_LG, -MCYCL_DS, -SER_TR, -VINA_MOD, -PREV_SUS)

vehicle80 <- vehicle80 %>%
  select(-CHAS_TR, -MCYCL_TY, -DR_TRAIN, -VIOL_CHG, -FLDCD_TR, -IMPACTS, -TOWAWAY, -VEH_CF1, -VEH_CF2, -DR_CF1, -DR_CF2, -DR_CF3, -IMPACT2, -WGTCD_TR, -VIN_LNGT, -VIN_WGT, -WHLBS_SH, -WHLBS_LG, -MCYCL_DS, -SER_TR, -VINA_MOD, -PREV_SUS)

vehicle85 <- vehicle85 %>%
  select(-HAZ_CARG, -VEH_MAN, -L_CL_VEH, -VIN_BT, -DR_TRAIN, -VIOL_CHG, -FLDCD_TR, -IMPACTS, -TOWAWAY, -VEH_CF1, -VEH_CF2, -DR_CF1, -DR_CF2, -DR_CF3, -IMPACT2, -WGTCD_TR, -VIN_LNGT, -VIN_WGT, -WHLBS_SH, -WHLBS_LG, -MCYCL_DS, -SER_TR, -VINA_MOD, -PREV_SUS)

vehicle90 <- vehicle90 %>%
  select(-HAZ_CARG, -VEH_MAN, -L_COMPL, -DR_ZIP, -VIN_BT, -VIOL_CHG, -FLDCD_TR, -IMPACTS, -TOWAWAY, -VEH_CF1, -VEH_CF2, -DR_CF1, -DR_CF2, -DR_CF3, -IMPACT2, -WGTCD_TR, -VIN_LNGT, -VIN_WGT, -WHLBS_SH, -WHLBS_LG, -MCYCL_DS, -SER_TR, -VINA_MOD, -PREV_SUS)

vehicle95 <- vehicle95 %>%
  select(-MONTH, -UNDERIDE, -VIN_11, -VIN_12, -HAZ_CARG, -VEH_MAN, -L_COMPL, -DR_ZIP, -VIN_BT, -OWNER, -V_CONFIG, -AXLES, -CARGO_BT, -AVOID, -CDL_STAT, -L_ENDORS, -VIOL_CHG, -FLDCD_TR, -IMPACTS, -TOWAWAY, -VEH_CF1, -VEH_CF2, -DR_CF1, -DR_CF2, -DR_CF3, -IMPACT2, -WGTCD_TR, -VIN_LNGT, -VIN_WGT, -WHLBS_SH, -WHLBS_LG, -MCYCL_DS, -SER_TR, -VINA_MOD, -PREV_SUS)

vehicle00 <- vehicle00 %>%
  select(-MONTH, -UNDERIDE, -VIN_11, -VIN_12, -HAZ_CARG, -VEH_MAN, -L_COMPL, -DR_ZIP, -VIN_BT, -OWNER, -V_CONFIG, -AXLES, -CARGO_BT, -AVOID, -CDL_STAT, -L_ENDORS, -BUS_USE, -GVWR, -MCARR_ID, -VIOLCHG1, -VIOLCHG2, -VIOLCHG3, -DR_CF4, -DR_WGT, -DR_HGT, -FLDCD_TR, -IMPACTS, -TOWAWAY, -VEH_CF1, -VEH_CF2, -DR_CF1, -DR_CF2, -DR_CF3, -IMPACT2, -WGTCD_TR, -VIN_LNGT, -VIN_WGT, -WHLBS_SH, -WHLBS_LG, -MCYCL_DS, -SER_TR, -VINA_MOD, -PREV_SUS)

vehicle05 <- vehicle05 %>%
  select(-MONTH, -UNDERIDE, -VIN_11, -VIN_12, -HAZ_CARG, -VEH_MAN, -L_COMPL, -DR_ZIP, -VIN_BT, -OWNER, -V_CONFIG, -AXLES, -CARGO_BT, -AVOID, -CDL_STAT, -L_ENDORS, -BUS_USE, -GVWR, -MCARR_ID, -VIOLCHG1, -VIOLCHG2, -VIOLCHG3, -DR_CF4, -DR_WGT, -DR_HGT, -SEQ1, -SEQ2, -SEQ3, -SEQ4, -SEQ5, -SEQ6, -UNITTYPE, -L_TYPE, -FLDCD_TR, -IMPACTS, -TOWAWAY, -VEH_CF1, -VEH_CF2, -DR_CF1, -DR_CF2, -DR_CF3, -IMPACT2, -WGTCD_TR, -VIN_LNGT, -VIN_WGT, -WHLBS_SH, -WHLBS_LG, -MCYCL_DS, -SER_TR, -VINA_MOD, -PREV_SUS)

vehicle10 <- vehicle10 %>%
  rename(OCUPANTS = NUMOCCS)

vehicle10 <- vehicle10 %>%
  select(-MONTH, -UNDERIDE, -VIN_11, -VIN_12, -L_COMPL, -DR_ZIP, -VIN_BT, -OWNER, -V_CONFIG, -CARGO_BT, -CDL_STAT, -L_ENDORS, -BUS_USE, -GVWR, -MCARR_ID, -DR_WGT, -DR_HGT, -UNITTYPE, -L_TYPE,
         -DAY, -HOUR, -MINUTE, -VINTYPE, -VINMAKE, -VINMODYR,
         -FUELCODE, -MCARR_I1, -MCARR_I2, -HAZ_INV, -HAZ_PLAC, -HAZ_ID,
         -HAZ_CNO, -HAZ_REL, -ROLINLOC, -TOWED, -VEH_SC1, -VEH_SC2,
         -SPEEDREL, -DR_SF1, -DR_SF2, -DR_SF3, -DR_SF4, -VTRAFWAY,
         -VNUM_LAN, -VSPD_LIM, -VALIGN, -VPROFILE, -VPAVETYP, -VSURCOND,
         -VTRAFCON, -VTCONT_F, -P_CRASH1, -P_CRASH2, -P_CRASH3, -PCRASH4,
         -PCRASH5, -ACC_TYPE, -IMPACT2, -WGTCD_TR, -VIN_LNGT, -VIN_WGT, -WHLBS_SH, -WHLBS_LG, -MCYCL_DS, -SER_TR, -VINA_MOD, -PREV_SUS)

vehicle15 <- vehicle15 %>%
  rename(OCUPANTS = NUMOCCS)

columns_15 <- colnames(vehicle15)
columns_00 <- colnames(vehicle00)

# Find the common columns
common_columns <- intersect(columns_15, columns_00)

# Select only the common columns from the first table
vehicle15 <- vehicle15[, common_columns]

vehicle20 <- vehicle20 %>%
  rename(OCUPANTS = NUMOCCS)

columns_20 <- colnames(vehicle20)

common_columns <- intersect(columns_20, columns_00)

vehicle20 <- vehicle20[, common_columns]


over_time <- rbind(vehicle75, vehicle80, vehicle85, vehicle90, vehicle95, vehicle00, vehicle05, vehicle10, vehicle15, vehicle20)

print(nrow(over_time))

library(ggplot2)

# Create a bar plot
ggplot(data = over_time, aes(x = factor(YEAR))) + 
  geom_bar() +
  labs(title = 'Entries per Year', x = 'Year', y = 'Number of Entries') +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

Link To Data: https://www.nhtsa.gov/file-downloads?p=nhtsa/downloads/FARS/



